---
title: "Final Rmarkdown Report"
author: "Corinne Hollyman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

FILL OUT

# BACKGROUND

Asthma is a chronic, or long-term condition that affects the airways of the lungs. Asthma causes inflammation of these passages, which in turn narrows the airways. This narrowing causes symptoms like shortness of breath, chest tightness, and coughing. While many consider asthma to be a childhood disease, asthma is also common in adults, affecting nearly 23 million US adults (Asthma in Adults | AAFA.org, 2024). Asthma triggers are things that can set off or worsen asthma symptoms, and include pollen, exercise, viral infections, or cold air (Keet et al., 2017). Air pollution, both indoor and outdoor, has also been linked to asthma symptoms (What Is Asthma? | NHLBI, NIH, 2024). While air pollution and asthma have been determined to be linked (Keet et al., 2017), it is not entirely clear if air pollution necessarily causes asthma, or if it simply causes or exacerbates symptoms in individuals already affected with asthma.


# STUDY QUESTION and HYPOTHESIS

## Questions  

Is there a correlation between the proportion of days with poor air quality and the percent of adults with current asthma?

## Hypothesis 

If poor air quality and asthma prevalence are linked, then states with more days with poor air quality will have more adults with asthma.

## Prediction

States with more days with recorded poor air quality will have higher prevalence of adults with asthma.

# METHODS

FILL OUT




## First Analysis- Scatterplot

This visualization plots the proportion of days with poor air quality on the x-axis, and the percentage of adults with current asthma on the y-axis. Each data point corresponds to a state in the US. The slope of the red line is the result of a linear regression model performed on the data.

```{r}
#importing data
my_data <- read.csv("Updated Final Data.csv")


#running a linear regression model on the data
regression_model <- lm(my_data$Percent.Adults.with.Current.Asthma ~ my_data$Percentage.of..poor.air.quality..days)


#creating scatterplot 
plot(my_data$Percentage.of..poor.air.quality..days, my_data$Percent.Adults.with.Current.Asthma,
     main = "Poor Air Quality Days vs Percent Adults with Current Asthma",
     xlab = "Percentage of days with \"poor\" air quality",
     ylab = "Percentage of adults with current asthma",
     pch = 16,
     col = "blue4")

#adding regression line to the plot
abline(regression_model, col="red") 

#adding a legend with the slope, p-value, and multiple r-squared values from the regession line
legend("topright",
       legend= c("Slope= -4.5919","p-value= 0.0497", "Multiple R-squared= 0.07785"),
       cex = 0.6,
       bty = "n")


```



## Second Analysis- Pearson Correlation Test

This test measures the correlation between the proportion of days with poor air quality and the percent of adults with current asthma in each state. The analysis calculates a Pearson Correlation value between -1 and 1. A high score indicates high similarity, while a score near 0 indicates no correlation (ADD SOURCE).

```{r}
#run correlation test
rce <- cor(my_data$Percentage.of..poor.air.quality..days, my_data$Percent.Adults.with.Current.Asthma)

rce #view test result
```



# DISCUSSION

## Interpretation of Scatterplot
FILL OUT 

## Interpretation of Second Analysis
FILL OUT 

# CONCLUSION
FILL OUT

# REFERENCES     

1. Asthma in adults | AAFA.org. (2024, October 7). Asthma & Allergy Foundation of America. https://aafa.org/asthma/living-with-asthma/asthma-in-adults/

2. Keet, C. A., Keller, J. P., & Peng, R. D. (2017). Long-Term Coarse Particulate Matter Exposure Is Associated with Asthma among Children in Medicaid. American Journal of Respiratory and Critical Care Medicine, 197(6), 737–746. https://doi.org/10.1164/rccm.201706-1267oc

3. What is asthma? | NHLBI, NIH. (2024, April 17). NHLBI, NIH. https://www.nhlbi.nih.gov/health/asthma

