---
title: "Final Rmarkdown Report"
author: "Corinne Hollyman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

FILL OUT

# BACKGROUND

FILL OUT 


# STUDY QUESTION and HYPOTHESIS

## Questions  

DRAFT: Is there a correlation between the proportion of days with poor air quality and the percent of adults with current asthma?

## Hypothesis 

FILL OUT

## Prediction
FILL OUT

# METHODS

FILL OUT

## First Analysis- Scatterplot
```{r}
my_data <- read.csv("Data for Final Rmarkdown Project.csv")

regression_model <- lm(my_data$Percent.Adults.with.Current.Asthma ~ my_data$Poor.Air.Quality.Days)

plot(my_data$Poor.Air.Quality.Days, my_data$Percent.Adults.with.Current.Asthma,
     ylim = c(8,13.5),
     main = "Title TBD",
     xlab = "Percentage of days with \"poor\" air quality",
     ylab = "Percentage of adults with current asthma",
     pch = 16,
     col = "blue4")

abline(regression_model, col="red") 


```



## Second Analysis- TBD

This can be ignored, I was just testing this.
```{r}
regression_model <- lm(my_data$Percent.Adults.with.Current.Asthma ~ my_data$Poor.Air.Quality.Days)

summary(regression_model)
```



# DISCUSSION

## Interpretation of Scatterplot
FILL OUT 

## Interpretation of Second Analysis
FILL OUT 

# CONCLUSION
FILL OUT

# REFERENCES     

FILL OUT

