---
title: "Warm-up mini-Report: Mosquito Blood Hosts in \n Salt Lake City, Utah"
author: "Corinne Hollyman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

This report investigates the species that mosquitoes in the Salt Lake City area are taking bloodmeals from, and the prevalence of mosquitoes infected with West Nile Virus. Through COI sequencing, the species each mosquito took a bloodmeal from was identified, as well as if each sample was infected with West Nile Virus. This data was then analyzed to examine the correlation between House Finch bloodmeals and viral infections. House Finches were found to have a positive correlation with West Nile Virus prevalence, indicating that House Finches are acting as amplifying hosts of WNV in Salt Lake City, UT. 

# BACKGROUND

Fill in some text here that provides background info on the WNV system, the blood meal DNA extractions, PCR, sequencing, etc. and the foundation for our question/hypothesis.

For example, we can use the viremia duration (Kumar et al., 2003) bar plot (make sure to reference sources!!!)  to illustrate the potential importance of house finches in WNV transmission and as the logical foundation for our hypothesis that house finches serve as amplifying hosts for WNV... and the prediction that locations with more house finches in our blood host analysis are also the same locations with higher positive tests for WNV in mosquito pools... 

NOTE: Examples of data you can plot for the background info at https://github.com/saarman/BIOL3070/

```{r viremia, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# Choose some colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

# STUDY QUESTION and HYPOTHESIS

## Questions  


Which bird species have the highest potential for transmission of West Nile Virus in Salt Lake City?

## Hypothesis 

House finches carry a high potential for transmission of West Nile Virus to humans via mosquitoes. 

## Prediction
If house finches are acting as amplifying hosts for West Nile Virus, then areas with more house finch bloodmeals will have higher levels of WNV in the pool of mosquitoes.

# METHODS

Over 300 mosquitoes were collected at various locations in Salt Lake City, UT. Polymerase Chain Reaction (PCR) was performed to amplify the amount of mitochondrial DNA present in the bloodmeals taken by the mosquitoes. After verifying the success of the PCR process with gel electrophoresis, the mitochondrial DNA was sequenced using MinION sequencing. The COI sequence for each bloodmeal was then analyzed using BLASTn at NCBI to identify the species each bloodmeal was taken from. Each sample was also analyzed to determine if West Nile Virus was present.  

## First Analysis- Horizontal Barplots
Horizontal plots:
```{r horiz-plot,fig.align='left',fig.width=7,fig.height=7, warning=FALSE}
## import counts_matrix: data.frame with column 'loc_positives' (0/1) and host columns 'host_*'
counts_matrix <- read.csv("./bloodmeal_plusWNV_for_BIOL3070.csv")

## 1) Identify host columns
host_cols <- grep("^host_", names(counts_matrix), value = TRUE)

if (length(host_cols) == 0) {
  stop("No columns matching '^host_' were found in counts_matrix.")
}

## 2) Ensure loc_positives is present and has both levels 0 and 1 where possible
counts_matrix$loc_positives <- factor(counts_matrix$loc_positives, levels = c(0, 1))

## 3) Aggregate host counts by loc_positives
agg <- stats::aggregate(
  counts_matrix[, host_cols, drop = FALSE],
  by = list(loc_positives = counts_matrix$loc_positives),
  FUN = function(x) sum(as.numeric(x), na.rm = TRUE)
)

## make sure both rows exist; if one is missing, add a zero row
need_levels <- setdiff(levels(counts_matrix$loc_positives), as.character(agg$loc_positives))
if (length(need_levels)) {
  zero_row <- as.list(rep(0, length(host_cols)))
  names(zero_row) <- host_cols
  for (lv in need_levels) {
    agg <- rbind(agg, c(lv, zero_row))
  }
  ## restore proper type
  agg$loc_positives <- factor(agg$loc_positives, levels = c("0","1"))
  ## coerce numeric host cols (they may have become character after rbind)
  for (hc in host_cols) agg[[hc]] <- as.numeric(agg[[hc]])
  agg <- agg[order(agg$loc_positives), , drop = FALSE]
}

## 4) Decide species order (overall abundance, descending)
overall <- colSums(agg[, host_cols, drop = FALSE], na.rm = TRUE)
host_order <- names(sort(overall, decreasing = TRUE))
species_labels <- rev(sub("^host_", "", host_order))  # nicer labels

## 5) Build count vectors for each panel in the SAME order
counts0 <- rev(as.numeric(agg[agg$loc_positives == 0, host_order, drop = TRUE]))
counts1 <- rev(as.numeric(agg[agg$loc_positives == 1, host_order, drop = TRUE]))

## 6) Colors: reuse your existing 'cols' if it exists and is long enough; otherwise generate
if (exists("cols") && length(cols) >= length(host_order)) {
  species_colors <- setNames(cols[seq_along(host_order)], species_labels)
} else {
  species_colors <- setNames(rainbow(length(host_order) + 10)[seq_along(host_order)], species_labels)
}

## 7) Shared x-limit for comparability
xmax <- max(c(counts0, counts1), na.rm = TRUE)
xmax <- if (is.finite(xmax)) xmax else 1
xlim_use <- c(0, xmax * 1.08)

## 8) Plot: two horizontal barplots with identical order and colors
op <- par(mfrow = c(1, 2),
          mar = c(4, 12, 3, 2),  # big left margin for species names
          xaxs = "i")           # a bit tighter axis padding

## Panel A: No WNV detected (loc_positives = 0)
barplot(height = counts0,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (-)",
        xlim = xlim_use)

## Panel B: WNV detected (loc_positives = 1)
barplot(height = counts1,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (+)",
        xlim = xlim_use)

par(op)

## Keep the colors mapping for reuse elsewhere
host_species_colors <- species_colors
```



## Second Analysis- Generalized Linear Model
```{r}
#glm with house-finch alone against positivity rate
glm2 <- glm(loc_rate ~ host_House_finch,
            data = counts_matrix)
summary(glm2)
```



# DISCUSSION

## Interpretation of Horizontal Barplots
The bar plot on the left demonstrates the distribution of species that bloodmeals were taken at locations that were negative for West Nile Virus. The bar plot on the right looks at the distribution of species at locations that were positive for West Nile Virus. In both plots, the top three species were House Finches, House Sparrows, and American Robins. However, at locations that were positive for West Nile Virus, the number of bloodmeals taken from House Finches was much higher, demonstrating a positive correlation between bloodmeals from House Finches and prevalence of West Nile Virus. 

## Interpretation of Generalized Linear Model
The generalized linear model looks at the relationship between the number of bloodmeals taken from House Finches and the number of mosquitoes infected with West Nile Virus per location. The estimate of this relationship is 0.0275, meaning that there is a slight positive correlation between the number of bloodmeals from House Finches and the number of infected mosquitoes. While this correlation doesn't appear strong, it is statistically significant, as it has a p-value of 4.54e-05, meaning that it is extremely unlikely that this correlation is due simply to chance. 
# CONCLUSION

# REFERENCES     

1. Komar N, Langevin S, Hinten S, Nemeth N, Edwards E, Hettler D, Davis B, Bowen R, Bunning M. Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerg Infect Dis. 2003 Mar;9(3):311-22. https://doi.org/10.3201/eid0903.020628

2. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.



